---
title: "Danish Kings week 10"
author: "Nanna Rolle og Aja Møller Jensen"
date: 'created on 7. Marts 2025, updated `r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, echo=TRUE}

#først starter jeg med at kører 'tidyverse' pakken, fordi man skal gøre dette hver gang åbner programmet
library(tidyverse)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:_______  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs

kings1 <- read.csv("data/DKMON.csv")
glimpse(kings1)

kings2 <- read_csv("data/DKMON.csv", na="NA")
glimpse(kings2)

kings3 <- read.csv2("data/DKMON.csv")
glimpse(kings3)

kings4 <- read_csv2("data/DKMON.csv")
glimpse(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object

- Funktion Kings2 og Kings4 er 'tidyverse'-funktion og det ser vi ved at der er underscore 'read_csv'. 
- funktion Kings1 og Kings2 er semikolon seperet fil, hvor Kings3 og Kings4 er kommaseperet fil.
- derfor skal jeg bruge Kings2, fordi min fil er semikolon seperert og jeg skal bruge 'tidyverse'funktionen, og det er Kings2 der kobler disse to. 

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

- her ser vi at det er 'tbl' og ikke 'dataframe' og vi vil gerne have 'tbl' fordi det er den gode form for uploading af et spreadsheet :), fordi man kan bruge tidyverse kodepsprog

3. How many columns does the object have when created with these different functions?

- Kings 3 og Kings 4 har 1 kolonne og Kings 1 og Kings 2 har 7 kolonner 

4. Show the dataset so that we can see how R interprets each column
- det kan man gøre med funktionen glimpse()

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

dkmon <- read_csv("data/DKMON.csv",na="NA")
glimpse(dkmon)

class(dkmon)
head(dkmon)
#her ser vi at det er 'tbl' og ikke 'dataframe' og vi vil gerne have 'tbl' fordi det er den gode form for uploading af et spreadsheet :), fordi man kan bruge tidyverse kodepsprog

#antal kolonner
ncol(dkmon)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE'
# mutate() to create new columns on the basis of two existing colums
##først laver vi en ny kollonne der hedder 'duration' som udregner regeringstiden
## derefter udregner vi regerings midt-året for hver regent

dkmonduration <- dkmon %>% 
  mutate(duration=Regerings_Slut - Regerings_Start_, na.rm=TRUE) %>% 
  mutate(Regerings_Midt_ = Regerings_Start_ + duration/2)

#og så laver vi en 'glimpse()' for at tjekke at der nu er en ny række der hedder 'Regerings_Midt_' og række der hedder 'duration', der viser gennemsnittet af længden af regeringsperioden
glimpse(dkmonduration)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
##gennemsnit
## vi beder R om at udregne gennemsnit (avg_duration), dette gøres med funktionen: mean(dkmonduration$duration,na.rm=TRUE). 
#$ = fortæller at vi kun vil bruge denne kollonne og 'na.rm=TRUE' fortæller R at den ikke skal medtage alle NA 
avg_duration <- mean(dkmonduration$duration,na.rm=TRUE)

#herefter kører vi koden'avg_duration' for at se svaret:
avg_duration #hermed får vi at gennemsnits længden for regeringstid er 19,66 år

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
# YOUR CODE
#her siger vi at R skal 'filter' de 'duration'  der er større end 'avg_duration'
long_reign_king <- dkmonduration %>% 
  filter(duration>avg_duration)

count(long_reign_king) #viser at 27 regenter har regeret længere end gennemsnittet på 19,66år

#for at vise de 27 regenter der har regeret længere end den gennemsnitlige regeringstid på 19,66 år bruger vi nedstående kode. Ved select --> siger vi at den skal tage kolonnen 'monark' og 'duration', samt at den skal vise alle 27 monarker.  
long_reign_king %>% 
  arrange(desc(duration))%>% 
  select(Monark, duration)%>% 
  print(n=27)


```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
#først laver vi et objekt der hedder 'largest_ruling_kings'
#desc --> sortere værdier i faldende rækkefølge
#med funktionen: 'slice_max(order_by=duration, n=3)' får vi R til kun at tage de tre regenter der har reageret længst
longest_ruling_kings <- dkmonduration %>% 
  arrange(desc(duration)) %>% 
  slice_max(order_by=duration, n=3)

longest_ruling_kings <- longest_ruling_kings %>% 
  mutate(Days=duration*365)

print(longest_ruling_kings)
#svar: Christian d. 4 (21900 dage), Dronning Magrethe d. 2 (18980 dage), Christian d. 9. (15695 dage)
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE

##For at plotte en graf
dkmonduration %>% 
  ggplot( aes(x=Regerings_Midt_, y=duration))+
  geom_point()+
  geom_smooth()+
  theme_classic()+
  xlab("Årstal")+
  ylab("Regeringstid i år")+
  ggtitle("Udvikling i regeringslængde for danske monarker")

## for at se de regenter der har regeret kortere end 10 år. Her ses det tydeligt at at der var flest(15) regenter der regerede under 10 år før 1340. 
short_reign_king <- dkmonduration %>% 
  filter(duration<10)

short_reign_king

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


